\proglang{Stan}: A Probabilistic Programming Language

### 2013-12-26 src/docs/papers/stan-language/article.tex
finished revisions of Stan paper

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Bob Carpenter
  \\
  Department of Statistics
  \\ 
  Columbia University
  \\ 
  1255 Amsterdam Avenue
  \\ 
  New York, NY 10027
  \\
  U.S.A.
  \\
  E-mail: \email{carp@stat.columbia.edu}
  \\ 
  URL: \url{http://mc-stan.org/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

@@

\section{TO-DO}

\begin{enumerate}
\item Convergence diagnostics?
\item more model examples?
\item description from C++ perspective?
\item what else do we want to make sure users don't miss?
\item optimization example?
\item future plans section?
\end{enumerate}

@@

The primitive types of \proglang{Stan} are \code{real} and \code{int},
which are used to represent continuous and integral values.  These
values are represented directly in \proglang{C++}.  Integer
expressions can be used anywhere a real value is required, but not
{\it vice-versa}.

@@

Higher-dimensional arrays are intrinsically arrays of arrays.  An
entry two-dimensional array \code{y} may be accessed as \code{y[1,2]}
or \code{y[1][2]}, with the expression \code{y[1]} denoting an array.%
%
\footnote{ Arrays are stored in row-major order and matrices in
  column-major order.}

@@

If necessary, an integer type will be promoted to a \code{real} value.
For example, multiplying an \code{int} by a \code{real} produces a
\code{real} result by promoting the \code{int} argument to a
\code{real}.  \proglang{Stan} also attempts to treat arrays as
covariant, meaning that arrays of of \code{int} values can often be
used where arrays of \code{real} values are specified.

@@

There is an implicitly defined variable \code{lp\_\_} (available in
the transformed parameters and model blocks) denoting the log
probability that will be returned by the log probability function.  A
sampling statement is nothing more than shorthand for incrementing the
log probability accumulator \code{lp\_\_}.  For example, if \code{beta}
is a parameter of type \code{real}, the sampling statement
%
\begin{Code}
beta ~ normal(0,1);
\end{Code}
%
has the exact same effect as the assignment statement
%
\begin{Code}
lp__ <- lp__ + normal_log(beta,0,1);
\end{Code}

Because computation is only up to a proportion, this sampling
statement in turn has the same effect as the direct implementation in
terms of basic arithmetic,
%
\begin{Code}
lp__ <- lp__ - 0.5 * beta * beta;
\end{Code}
%
If \code{beta} is of type \code{vector}, replace \code{beta * beta}
with \code{beta' * beta}.  Distributions whose probability functions
are not built directly into \proglang{Stan} can be implemented
directly in this fashion.

@@

Following the convention of \proglang{C++}, statements are separated
with semicolons in \proglang{Stan} so that the content of whitespace
(outside of comments) is irrelevant.  This is in contrast to
\proglang{BUGS} and \proglang{JAGS} in which carriage returns are
special and may indicate the end of a statement.

@@

Following the usual convention in mathematics, matrix and array
indexing uses the usual square brackets (\code{[ ]}) operator, and
begins indexing from 1.  If \code{Sigma} is a matrix, then
\code{Sigma[n]} is the \code{n}-th row of \code{Sigma}.

@@

All of the univariate%
%
\footnote{We are in the process of vectorizing the multivariate
  probability functions, but they are not all available in
  \proglang{Stan} 2.0.  We are also in the process of vectorizing
  all of the special functions.}
%
probability functions in \proglang{Stan} are vectorized so that they
accept arrays or vectors of arguments.  For example, although the
basic signature of the probability function
\code{normal\_log(y,mu,sigma)} involves real \code{y}, \code{mu} and
\code{sigma}, it supports calls in which any any or all of \code{y},
\code{mu} and \code{sigma} contain more than one element. A typical
use case would be for linear regression, such as \code{normal\_log(y,X
  * beta,sigma)}, where \code{y} is a vector of observed data,
\code{X} is a predictor matrix, \code{beta} is a coefficient vector,
and \code{sigma} is a real value for the noise scale.

The advantage of using vectorization is twofold.  First, the models
are more concise and closer to mathematical notation.  Second, the
vectorized versions are much faster.  They reduce the number of times
expensive operations need to be evaluated and also reduce the number
of virtual function calls required in the compiled \proglang{C++}
executable for calculating gradients and higher-order derivatives.
In the example above, \code{normal\_log(y,X *
  beta,sigma)}, the logarithm of \code{sigma} need only be computed
once; if \code{y} is an $N$-vector, it also reduces the number of
virtual function calls from $N$ to 1.

@@

The MCMC samplers provided include Euclidean Manifold Hamiltonian
Monte Carlo (EM-HMC) \citep{DuaneEtAl:1987, Neal:1994, Neal:2011} and
the No-U-Turn sampler (NUTS) \citep{HoffmanGelman:2011}.  Both
versions of HMC allow estimation or specification of unit, diagonal,
or full mass matrices.  NUTS, the default sampler for \proglang{Stan},
automatically adapts the number of leapfrog steps, eliminating the
need for user-specified tuning parameters.  Both algorithms take
advantage of curvature information, namely the gradient, to generate
coherent motion through the posterior that dramatically reduces the
autocorrelation of the resulting transitions.

\proglang{Stan} will soon have an implementation of Riemannian Manifold
Hamiltonian Monte Carlo (RM-HMC) \citep{GirolamiCalderhead:2011},
using the SoftAbs metric to robustly incorporate posterior curvature
into the Hamiltonian simulation \cite{Betancourt:2012}, although
\proglang{Stan}'s MCMC framework makes it
straightforward to incorporate other metrics.  This
implementation will also generalize NUTS to Riemannian manifolds
\citep{Betancourt:2013}.  Both support adapting step sizes during
warmup.  RM-HMC uses first, second, and third order derivatives 
to adapt to the local curvature of the posterior.

@@

\proglang{Stan} supports random-walk Metropolis-Hastings samplers with
multivariate normal proposals, the primary purpose of which is to
provide baselines against which to measure more efficient approaches
to sampling.  The covariance matrix used can be adapted during warmup,
assuming either equal diagonal covariance, diagonal covariance, or a
full covariance matrix.  Metropolis-Hastings does not require any
derivatives of the log probability function, but the lack of gradient
information induces random walk behavior.

@@

Two flavors of ensemble samplers are also in the works: an affine invariant
ensemble sampler \citep{GoodmanWeare:2010} and a differential
evolution sampler \cite{TerBraak:2006}.  Ensemble samplers do not
require derivative information, but typically require at least as many
chains to be run as the number of dimensions in order to have full
rank search of the posterior.

@@

The default optimizer in \proglang{Stan} is the
Broyden-Fletcher-Goldfarb-Shannon-Boyd (BFGS) optimizer.  BFGS is a
quasi-Newton optimizer that evaluates gradients directly, then uses
the gradients to update an approximation to the Hessian; see
\citep{NocedalWright:2006} for details.  Plans are in the works to
also include the more involved, but more scalable limited-memory BFGS
(L-BFGS) scheme.

@@

Additionally, \proglang{Stan} implements a crude version of Nesterov's
accelerated gradient optimizer \cite{Nesterov:1983} which combines
gradient updates with a momentum-like update to speed up convergence.

@@

This includes extensive unit tests for low-level \proglang{C++} code.
Unit tests are implemented using the \pkg{googletest} framework
\citep{GoogleTest:2011}.  The probability functions and command-line
invocations are complex enough that programs are used to automatically
generate test code for \pkg{googletest}.

>>>>

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Bob Carpenter
  \\
  Department of Statistics
  \\ 
  Columbia University
  \\ 
  1255 Amsterdam Avenue
  \\ 
  New York, NY 10027
  \\
  U.S.A.
  \\
  E-mail: \email{carp@alias-i.com}
  \\ 
  URL: \url{http://mc-stan.org/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

@@

@@

The primitive types of \proglang{Stan} are \code{real} and \code{int},
which are used to represent continuous and integer values.  These
values are represented directly in \proglang{C++} as types
\code{double} and \code{int}.  Integer expressions can be used
anywhere a real value is required, but not {\it vice-versa}.

@@

Higher-dimensional arrays are intrinsically arrays of arrays.  An
entry two-dimensional array \code{y} may be accessed as \code{y[1,2]}
or \code{y[1][2]}, with the expression \code{y[1]} denoting an array.%
%
\footnote{Arrays are stored in row-major order and matrices in
  column-major order.}

Unlike integers, which may be used where real values are required,
arrays of integers may not be used where real arrays are required.%
%
\footnote{In the language of type theory, \proglang{Stan} arrays are
  not covariant.  This follows the behavior of both arrays and
  standard library containers in \proglang{C++}.}

@@

If necessary, an integer type will be promoted to a \code{real} value.
For example, multiplying an \code{int} by a \code{real} produces a
\code{real} result by promoting the \code{int} argument to a
\code{real}.

@@

There is an implicitly defined variable \code{lp\_\_} (available in 
the transformed parameters and model blocks) denoting the log
probability that will be returned by the log probability function.  A
sampling statement is nothing more than shorthand for incrementing the
log probability accumulator \code{lp\_\_}.  For example, if \code{beta}
is a parameter of type \code{real}, the sampling statement
%
\begin{Code}
beta ~ normal(0,1);
\end{Code}
%
has the exact same effect as the assignment statement
%
\begin{Code}
increment_log_prob(normal_log(beta,0,1));
\end{Code}
%
The sampling statement distribution \code{normal} is shorthand for the
function \code{normal\_log}, and both the left-hand side variate and
right-hand side parameters become arguments.

Because computation is only up to a proportion, this sampling
statement in turn has the same effect as the direct implementation in
terms of basic arithmetic,
%
\begin{Code}
increment_log_prob(-0.5 * beta * beta);
\end{Code}
%
If \code{beta} is of type \code{vector}, replace \code{beta * beta}
with \code{beta' * beta}.  Distributions whose probability functions
are not built directly into \proglang{Stan} can be implemented
directly in this fashion.

@@

Following the convention of \proglang{C++}, statements are separated
with semicolons in \proglang{Stan} so that the content of whitespace
(outside of comments) is irrelevant.  This is in contrast to
\proglang{BUGS} and \proglang{R}, in which carriage returns are
special and may indicate the end of a statement.

@@

Following the usual convention in mathematics, matrix and array
indexing uses the usual square brackets (\code{[ ]}) operator, and
begins indexing from 1.  For example, if \code{Sigma} is a matrix,
then \code{Sigma[m,n]} is the entry at row \code{m} and column
\code{n}.  \proglang{Stan} also allows slices of data structures to be
returned, so that \code{Sigma[m]} is row \code{m} of the matrix
\code{Sigma}.

@@

All of the univariate%
%
\footnote{We are in the process of vectorizing the multivariate
  probability functions, but they are not all available in
  \proglang{Stan} 2.0.  We are also in the process of vectorizing
  all of the special functions.}
%
probability functions in \proglang{Stan} are vectorized so that they
accept arrays or vectors of arguments.  For example, although the
basic signature of the probability function
\code{normal\_log(y,mu,sigma)} involves real \code{y}, \code{mu} and
\code{sigma}, it supports calls in which any any or all of \code{y},
\code{mu} and \code{sigma} contain more than one element. A typical
use case would be for linear regression, such as \code{y ~ normal(X
  * beta,sigma)}, where \code{y} is a vector of observed data,
\code{X} is a predictor matrix, \code{beta} is a coefficient vector,
and \code{sigma} is a real value for the noise scale.

The advantage of using vectorization is twofold.  First, the models
are more concise and closer to mathematical notation.  Second, the
vectorized versions are much faster.  They reduce the number of times
expensive operations need to be evaluated and also reduce the number
of virtual function calls required in the compiled \proglang{C++}
executable for calculating gradients and higher-order derivatives.
In the example above, \code{y ~ normal(X *
  beta,sigma)}, the logarithm of \code{sigma} need only be computed
once; if \code{y} is an $N$-vector, it also reduces the number of
virtual function calls from $N$ to 1.

@@

The MCMC samplers provided include Euclidean Manifold Hamiltonian
Monte Carlo (EHMC, or just HMC) \citep{DuaneEtAl:1987, Neal:1994,
  Neal:2011} and the No-U-Turn sampler (NUTS)
\citep{HoffmanGelman:2011}.  Both versions of HMC allow estimation or
specification of unit, diagonal, or full mass matrices.  NUTS, the
default sampler for \proglang{Stan}, automatically adapts the number
of leapfrog steps, eliminating the need for user-specified tuning
parameters.  Both algorithms take advantage of gradient information in
the log probability function to generate coherent motion through the
posterior that dramatically reduces the autocorrelation of the
resulting transitions.

\proglang{Stan} will soon have an implementation of Riemannian
Manifold Hamiltonian Monte Carlo (RHMC)
\citep{GirolamiCalderhead:2011}, using the SoftAbs metric to robustly
incorporate posterior curvature into the Hamiltonian simulation
\cite{Betancourt:2012}.  \proglang{Stan}'s MCMC framework also makes
it straightforward to incorporate other metrics.  This implementation
will also generalize NUTS to Riemannian manifolds
\citep{Betancourt:2013}.  Both support adapting step sizes during
warmup.  RHMC with SoftAbs uses first, second, and third order
derivatives to adapt to the local curvature of the posterior.

@@

Random-walk Metropolis-Hastings samplers with multivariate normal
proposals have been prototyped, but not yet integrated into the
released versions of \proglang{Stan}.  The primary purpose of
implementing Metropolis-Hastings is to provide baselines against which
to measure more efficient approaches to sampling.  The covariance
matrix used can be adapted during warmup, assuming either equal
diagonal covariance, diagonal covariance, or a full covariance matrix.
Metropolis-Hastings does not require any derivatives of the log
probability function, but the lack of gradient information induces
random walk behavior, leading to slow mixing.

@@

Two flavors of ensemble samplers are also in the works: an affine invariant
ensemble sampler \citep{GoodmanWeare:2010} and a differential
evolution sampler \cite{TerBraak:2006}.  Ensemble samplers do not
require derivative information, but typically require at least as many
chains to be run as the number of dimensions in order to have full
rank search of the posterior.  They also require slightly different
management of output due to a single sampler instance producing
multiple correlated chains of draws.

@@

The default optimizer in \proglang{Stan} is the
Broyden-Fletcher-Goldfarb-Shannon-Boyd (BFGS) optimizer.  BFGS is a
quasi-Newton optimizer that evaluates gradients directly, then uses
the gradients to update an approximation to the Hessian.  Plans are in
the works to also include the more involved, but more scalable
limited-memory BFGS (L-BFGS) scheme.  \citep{NocedalWright:2006} cover
both BFGS and L-BFGS samplers.

@@

Additionally, \proglang{Stan} implements a crude version of Nesterov's
accelerated gradient optimizer \cite{Nesterov:1983}, which combines
gradient updates with a momentum-like update to hasten convergence.

@@

\proglang{Stan} includes extensive unit tests for low-level \proglang{C++} code.
Unit tests are implemented using the \pkg{googletest} framework
\citep{GoogleTest:2011}.  The probability functions and command-line
invocations are complex enough that programs are used to automatically
generate test code for \pkg{googletest}.
